Dessicated diversification: how aridification slowed down speciation of Australian blind snakes

### 2023-01-19 manuscript.Rmd
incorporated text from Alex Skeels and made changes in response to comments. Still need to explain why increase aridity decrease diversification

---
title: ""
date: "`r Sys.Date()`"
author: 
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [C:/Users/ST/Documents/repo/AusARG_blindsnake/custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
# biblio-style: "apalike"
csl: jbiogeography.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

Possible titles:
- Diversification rates of Australian blind snakes slows down as the continent became more dry up.
- Reconstructed paleoaridity correlate with decreased diversification rates in Australian blind snakes.
- Drying up speciation: aridification of Australia slowed down diversification of Australian blind snakes
- Dull when dry: increased aridity correlate with decrease diversification in blind snakes

**Abstract**

**Aim**: Shifts in diversification rates of Australian flora and fauna has been associated with aridification, but the relationship between diversification rates and aridity ifself has never been quantified. We investigate how paleoenvironmental changes have influenced the diversification rates of the mostly endemic Australian blind snakes genus, Anilios. We explored how changes in temperature and aridity during the Neogene have influenced the diversification of the Australian blind snake and we further investigate whether diversification rates differ between arid-adapted lineages vs mesic-adapted lineages.

@@

**Materials and Methods**: We infer the ancestral biome of Australian blind snakes using BioGeoBEARS. We reconstruct the paleoenvironments of Australia during the Neogene using the gen3sis engine. We fit several birth-death models and estimate diversification rates under paleoenvironmental conditions using RPANDA. We further compare diversification rates between arid-adapted lineages versus mesic-adapted lineages using ClaDS and GeoHiSSE.

**Results**: Results from our BioGeoBEARS analyses indicate that Australian blind snakes have tropical grassland origins. We found that local Australian paleoenvironments provide better fit than global paleoenvironments. Diversification rates of blind snakes decreased with increasing aridity. We found no difference in diversification rates between arid- and mesic-adapted lineages.

@@

**Keywords**
Scolecophidia, Diversification, gen3sis, BioGeoBEARS, GeoHiSSE

@@

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo and mode with which different groups diversified. The aridification of continental Australia began around 20 million years ago, with pockets of arid zone developing and subsequent extreme warming and cooling periods in the last ~5 million years creating what is now known as the "outback" [@byrneBirth2008]. During this period, arid-adapted vegetation expanded and mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to have involved the rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017].Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia's terrestrial reptiles, termites, and ants [@byrneBirth2008].

Squamate reptiles (lizards and snakes) represent the largest vertebrate diversity in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which colonised the continent from elsewhere and have diversified in the last 20 – 6 MYRS [@brennanMass2017;@oliverPhylogenetic2017].

Some clades, like elapid snakes and skinks, showed dramatic morphological and ecological differentitation [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Other clades, like some geckos and blind snakes, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. The species richness and distribution of squamate reptiles across different biomes makes them a good group to test hypotheses about the tempo and mode of diversification in Australia.

Roughly a quarter of the snake species diversity in Australia is represented by blind snakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021] (**Tiatragul et al. 2023 lumped on**). All currently described species of _Anilios_ are found in Australia except for _A. eryncinus_, which is endemic to New Guinea. Blind snakes are morphologically conserved and often difficult to distinguish, and extensive phylogenetic studies and several efforts to revisit specimens have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single colonisation event by a mesic-adapted ancestor from Asia between 39 – 19 MYA [@vidalBlindsnake2010;@marinTracing2013]. However, despite representing a substantial portion of snake biodiversity, there has been few investigations into the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted and can be found in every biome on mainland Australia.

<!-- Environmental correlates of diversification -->
<!-- agamid lizards [@harmonTempo2003], passerine birds [@ricklefsGlobal2006], amphibians [@rollandContribution2019], rosid plants [@sunRecent2020]).  -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], elevation [@esquerreHow2019;@boschmanMountain2022], biological interactions [@ramirezAsynchronous2011;@druryEstimating2016;@brennanPhylogenomics2021], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014], frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and plants (reviewed in @crispHow2013).

In reptiles, variation in climate has been proposed to influence diversification rates. For instance, warmer temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Second, spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Third, change in climate (e.g. glaciation and aridification) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Fourth, ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, particularly mesic-adapted lineages that have colonised Australia from elsewhere.
 <!---which have been documented in several arid-adapted lineages with deep divergences [@byrneBirth2008;@pepperGenetic2011;@marinTracing2013]. --->

The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019]. Traditionally, these temperature data are reconstructed (interpolated) based on conversion of global concentration of oxygen isotopes [@condamineMacroevolutionary2013]. The relationship between diversification rates and aridity, however, have been attributed to some proxies such as the timing of aridification or global temperature rather than quantifying the extent of local aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015]. 
To our knowledge no study have assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent development in geographic-based methods such as `gen3sis` modelling engine [@hagenGen3sis2021] enables us to simulate paleotemperature and paleoaridity data specific to Australia in the time span of interest. These data sets allow us to empirically test the influence of localised ecological processes on diversification rates over time.

The aim of this study is to infer the ancestral biome of Australian blind snakes radiation and investigate how paleoenvironmental changes have influenced the rates of diversification within this group. To do this, we fit several birth-death models and estimate diversification rates under different conditions. We further compare diversification rates between arid-adapted lineages versus non arid-adapted lineages.

@@

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employed a time-calibrated phylogenomic tree of Australian blind snakes with outgroups (Tiatragul et al. 2023). This tree includes 38 operational taxonomic units (OTUs) representing 36/47 described species of _Anilios_. However, our phylogenetic data suggest the existence of at least three genetically distinct undescribed species, so we adjusted our taxon sampling fraction to 38/50 species. The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and 2 '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (see Tiatragul et al. 2023). For ancestral biome estimation analyses, we also included _Ramphotyphlops multilineatus_, _Sundatyphlops polygrammicus_, and _Acutotyphlops subocularis_ as outgroup taxa.

@@

To investigate the biogeographic history of blind snakes, we used the R package BioGeoBEARS v.1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal, extinction, or switching to adjacent biomes) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
<!--The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range in which each species is found, and a historical biogeographic model.  -->
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

We applied the phylogeny and distribution information to widely used historical biogeographical methods, including likelihood implementations of Dispersal-Vicariance Analysis(DIVALIKE) [@ronquistDispersalVicariance1997], Dispersal-Extinction-Cladogenesis (DEC) [@reeMaximum2008], and BayArea (BAYAREALIKE) [@landisBayesian2013] in BioGeoBEARS [@matzkeBioGeoBEARS2013]. We included two additional free parameters that we thought were relevant. The first parameter, *j*, assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter, _x_, multiplies the probability of dispersal by distance to the power of _x_ [@vandamEvaluating2016] (see Appendix-method-biogeobears).
We used likelihood ratio tests to determine if the additional free parameter was a justified improvement over the simpler base models, and selected the model that best fitted the data based on corrected Akaike Information Criterion (AICc) scores. Based on the best model, we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and transition between biomes, and to estimate the frequency and direction of dispersal events across biomes [@matzkeBioGeoBEARS2013].

## Paleo environment reconstruction
<!-- Consult Alex Skeels -->
To estimate paleoclimate, we used three paleographic digital elevation models (DEMs) from @scoteseAtlas2021, @straumeGlobal2020, and [VALDES citation]. Each DEM provides paleotopography at a resolution of 1&deg; x 1&deg; and 5 million years intervals. We projected the paleoenvironmental reconstructions into an equal-area Behrmann projection of 110km x 110km, following the methods described in @hagenGen3sis2021. From the projected data, we extracted temperature and aridity values across all grid cells that fall within Sahul by manually drawing a polygon that encompassed the continent from 40 MYA to the present day.
Based on the ____ value, we use XX as the threshold for arid. Thus aridity index is ...

@@

To evaluate how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the fit_bd and fit_env functions in the R package RPANDA [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with paleoenvironment as described above. We also fit the model to widely use mean global paleotemperature data set [@condamineMacroevolutionary2013]. To account for missing taxa, we used the fractioning function (f = 38/50) in RPANDA. We then assessed which diversification model best fit the data based on AICc scores.

@@

The four SSE models differed in whether they allowed diversification rates to be dependent on geographic states (range-dependent) or independent of them (range-independent), and whether they allowed an unmeasured trait ('hidden states') to influence diversification rates [@beaulieuDetecting2016;@caetanoHidden2018]. 
Model 1 assumed equal diversification rates regardless of geographic state. Model 2 allowed geographic range to affect diversification rates.
Models 3 and 4 were based on models 1 and 2, but they also allowed unmeasured traits ('hidden states') to contribute to diversification rates. By incorporating hidden states, we could assess whether geographic range or an unmeasured trait contributed more to the observed diversification. For each model, we assumed uniform taxon sampling for 89% of arid lineages, 71% of non-arid lineages, and 83% of widespread lineages, based on the number of species in the phylogeny and all currently described species of _Anilios_. We then calculated AIC to identify the best-fitting models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters across all models according to their AIC weights.

@@

The largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events), which makes up  43.6% compared outward dispersals from the arid biome (9.7 ± 3.2 events) of 26% (Fig. \@ref(fig:fig-biogeobears)c-d). Together the tropical grassland biome and the arid biome contributes the largest source of diversification accounting for 69.6% of outward transitions. On the other hand, the largest sink for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c-d).

Based on current distributions from ALA, most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and mediterranean biome (n = 10). We identified divergent biogeographic transitions between two major groups of species, which we will refer to as *Group 1* (n = 21 lineages, not including _A. proximus_) and *Group 2* (n = 14 lineages, not including _A. silvia_) (Fig. \@ref(fig:fig-biogeobears)a). The common ancestor of *Group 1* remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of *Group 2* first transitioned into the arid biome, and then 7/14 lineages expanded to the mediterranean and temperate forest biomes, and/or transitioned back into the arid biome.

@@

Paleotemperature estimates from gen3sis for Australia differed between the three DEMs (Fig. \@ref(fig:fig-paleoclimate-compare)b). However, all three paleotemperature models indicate Australia was hotter than the average global temperature (Fig. \@ref(fig:fig-paleoclimate-compare)a). Paleoclimate in Australia were also relatively constant compared to global temperature until the Plio-Pleistocene, a period characterised by fluctuations corresponding to the glacial cycles. 
Reconstructions of paleoaridity differed slightly between the two methods (Fig. \@ref(fig:fig-paleoclimate-compare)b). Both aridity index shows increase in aridity post 10 MYA, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021].

>>>>

---
title: "Dessicated diversification: how aridification slowed down speciation of Australian blind snakes"
date: "`r Sys.Date()`"
author: "Sarin Tiatragul^1^; Alexander Skeels^2,3^; J. Scott Keogh^1^"
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [C:/Users/ST/Documents/repo/AusARG_blindsnake/custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
# biblio-style: "apalike"
csl: jbiogeography.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

### Possible titles:

- Diversification rates of Australian blind snakes slows down as the continent became more dry up.

- Reconstructed paleoaridity correlate with decreased diversification rates in Australian blind snakes.

- Drying up speciation: aridification of Australia slowed down diversification of Australian blind snakes

- Dull when dry: increased aridity correlate with decrease diversification in blind snakes

**Address**: Division of Ecology & Evolution, Research School of Biology, The Australian National University, Canberra 2601, ACT, Australia

1 Division of Ecology & Evolution, Research School of Biology, The Australian National University, Canberra 2601, ACT, Australia

2 Department of Environmental Systems Sciences, Landscape Ecology, Institute of Terrestrial Ecosystems, ETH Zürich, Zurich 8092, Switzerland;

3 Swiss Federal Institute for Forest, Snow and Landscape Research WSL, Birmensdorf 8903, Switzerland;

**Corresponding author**: Sarin Tiatragul (sarin.tiatragul@anu.edu.au)

### J. Biogeography format::

**Aim**: Shifts in diversification rates of Australian flora and fauna has been associated with the process of aridification, but the relationship between diversification rates and aridity has never been quantified. We investigate how paleoenvironmental changes have influenced the diversification rates of the mostly endemic Australian blind snakes genus, Anilios. We explored how changes in temperature and aridity during the Neogene have influenced the diversification of the Australian blind snake and we further investigate whether diversification rates differ between arid-adapted lineages vs mesic-adapted lineages.

@@

**Materials and Methods**: We estimated the historical biogeography of blind snakes using BioGeoBEARS. 
We synthesised multiple approaches to reconstruct paleotemperature and paleoaridity of Australia during the Neogene. We fit several birth-death models and estimate diversification rates under paleoenvironmental conditions using RPANDA. We further compare diversification rates between arid-adapted lineages versus mesic-adapted lineages using ClaDS and GeoHiSSE.

**Results**: Ancestral area estimation indicated Australian blind snakes have tropical grassland origins. We found that local Australian paleoenvironments provided a better explanation for diversification rate variation than global paleotemperature. Speciically, diversification rates of blind snakes decreased with increasing aridity. We found no difference in diversification rates between arid- and mesic-adapted lineages.

@@

**Keywords**
Scolecophidia, Diversification, paleoenvironment, BioGeoBEARS, GeoHiSSE

@@

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo and mode with which different groups diversified. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annums) with the development of pockets of aridity [@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, arid-adapted vegetation expanded and mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to have promoted the rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;
@owenHow2017].

<!-- Environmental correlates of diversification -->
A variety of ecological correlates have been associated with diversification processes of a clade. Among others, studies on the influence of climate (e.g. temperature, precipitation, and latitude) [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], elevation [@esquerreHow2019;@boschmanMountain2022], biological interactions [@ramirezAsynchronous2011;@druryEstimating2016;@brennanPhylogenomics2021], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017] and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014] have been documented. In Australia, shifts in diversification rates owing to changes in temperature and aridity has been reported in skinks [@raboskyExceptional2007], geckos [@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014], frogs [@catulloAridification2014;@vidal-garciaConvergent2015], saltbush [@kadereitMolecular2010], *Acacia* [@millerComparative2013], insects [@owenHow2017;@toussaintUnveiling2015], and plants [@crispHow2013].

<!-- paragraph 5 -->
In reptiles, variation in climate has been proposed to influence diversification rates. For instance, warmer temperature may increase resource availability, which in turn increase ecological niche in general [@clarkeClimate2006]. Spatial and temporal heterogeneity in temperature may lead to climatic niche divergence, such as local adaptations in different geographic areas [@pepperGeodiversity2013], and change in activity time of prey/predators [@santosTemperature2011;@lawsonLatitudinal2014]. Change in climate (e.g. glaciation and aridification) may cause fragmentation of continuous habitats leading to isolated populations, which may lead to divergence [@oliverRocks2019]. Ectotherms are subjected to higher metabolic rates in warm climates (i.e. Kleiber's law), which may increase mutation rates and ultimately greater speciation rates [@allenKinetic2006;@makarievaMean2008]. Finally, aridification of Australia may offer new ecological niche for species to exploit, particularly mesic-adapted lineages that have colonised Australia from elsewhere.
 <!---which have been documented in several arid-adapted lineages with deep divergences [@byrneBirth2008;@pepperGenetic2011;@marinTracing2013]. --->

<!-- paragraph 3 -->
Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia's terrestrial reptiles, termites, and ants [@byrneBirth2008]. Squamate reptiles (lizards and snakes) are the largest terrestrial vertebrate group in Australia with over 1000 species (~800 lizards and ~200 snakes), many of which colonised the continent from elsewhere and diversified in the last 20 – 6 Ma [@brennanMass2017;@oliverPhylogenetic2017;@brennanPhylogenomics2021;@esquerrePhylogenomics2020]. Some clades, like elapid snakes and skinks, showed dramatic morphological and ecological differentitation [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Other clades, like some geckos and blind snakes, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. The species richness and distribution of squamate reptiles across different biomes makes them a good group to test hypotheses about the tempo and mode of diversification in Australia.

<!-- paragraph 4 -->
Roughly a quarter of the snake species diversity in Australia is represented by blind snakes of the genus _Anilios_, which comprises **47** described species [@uetzQuarter2021] (**Tiatragul et al. under review**). All currently described species of _Anilios_ are fossorial and are found in Australia except for _A. eryncinus_, which is endemic to New Guinea. Blind snakes are morphologically conserved, species often difficult to distinguish, and extensive phylogenetic studies and several efforts to revisit specimens have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single colonisation event by a mesic-adapted ancestor from Asia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. Despite representing a substantial portion of snake biodiversity, we know little about the macroevolutionary dynamics of blind snakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic ecological changes because they are geographically restricted and can be found in every biome on mainland Australia.

<!-- paragraph 6 -->
The influence of temperature on diversification rates has been empirically tested with birth-death models [@condamineMacroevolutionary2013;@kergoatOpposite2018;@condamineClimate2019;@rollandContribution2019]. Traditionally, these temperature data are reconstructed (interpolated) based on conversion of global concentration of oxygen isotopes [@condamineMacroevolutionary2013]. The relationship between diversification rates and aridity, however, have been attributed to some proxies such as the timing of aridification or global temperature rather than quantifying the extent of local aridity itself [@dornStrange2014;@deepakAridification2018;@catulloAridification2014;@fujitaDiversification2010;@pepperGenetic2011;@pepperPalaeoclimate2011;@oliverIndependent2014;@oliverContrasting2014;@owenHow2017;@crispHow2013;@toussaintUnveiling2015].

<!-- Paragraph 7 -->
To our knowledge no study has assessed aridity as a continuous variable in an environment-dependent birth-death model framework, probably due to the difficulty of quantifying paleoaridity. However, recent developments in spatially-explicit paleoenvironmental reconstructions from the lithological and fossil record as well as global circulation models (GCMs) now enables us to estimate paleotemperature and paleoaridity data specific to Australia in the time span of interest. These data sets for the first time enables us to empirically test the effects of localised environmental changes on diversification rates over time.

<!-- Paragraph 8 -->
The aim of this study is to infer the historical biogeography of the Australian blind snake radiation and investigate how paleoenvironmental changes have influenced the rates of diversification within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimate diversification rates under these models. We further compare diversification rates between arid-adapted lineages versus non arid-adapted lineages.

@@

## Phylogenetic hypothesis and taxon sampling
<!-- Based on our phylogenetic tree -->
We employed the most comprehensive time-calibrated phylogenomic tree of Australian blind snakes with outgroups (Tiatragul et al. under review). This tree includes 38 operational taxonomic units (OTUs) representing 36/47 described species of _Anilios_. The remaining 11 species could not be included because no molecular data or tissues samples are available. However, our phylogenetic data suggest the existence of at least three genetically distinct undescribed species, so we adjusted our taxon sampling fraction to 38/50 species. The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and 2 '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (see Tiatragul et al. under review). For ancestral biome estimation analyses, we also included _Ramphotyphlops multilineatus_, _Sundatyphlops polygrammicus_, and _Acutotyphlops subocularis_ as outgroup taxa.

@@

To investigate the biogeographic history of blind snakes, we used the R package BioGeoBEARS v.1.1.1 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
<!--The fundamental inputs for the software include a time-calibrated phylogeny, information about the biome/range in which each species is found, and a historical biogeographic model.  -->
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

Using the phylogeny and distribution data, we fitted three widely used historical biogeographical methods, which vary in the kinds of cladogenetic and anagenetic events they model, including likelihood implementations of Dispersal-Vicariance Analysis(DIVALIKE) [@ronquistDispersalVicariance1997], Dispersal-Extinction-Cladogenesis (DEC) [@reeMaximum2008], and BayArea (BAYAREALIKE) [@landisBayesian2013] in the R package BioGeoBEARS [@matzkeBioGeoBEARS2013]. 
We also fitted these three models, with two additional free parameters that we thought were relevant, both singly and together, for a total of 12 alternative models. The first additional parameter, *j*, assigns per-event weight to founder-events [@matzkeModel2014], and the second parameter, _x_, multiplies the probability of dispersal by distance to the power of _x_ [@vandamEvaluating2016] (see Appendix-method-biogeobears).
We used likelihood ratio tests to determine if the additional free parameter was a justified improvement over the simpler base models, and selected the model that best fitted the data based on corrected Akaike Information Criterion (AICc) scores. Based on the best model, we ran an additional 50 biogeographic stochastic maps to account for uncertainty in ancestral biome estimation and transition between biomes, and to estimate the frequency and direction of dispersal events across biomes [@matzkeBioGeoBEARS2013].

## Paleo environment reconstruction
<!-- Consult Alex Skeels -->

To estimate paleoclimate in Australia from 40 Ma to the present-day, we used three two alternative approaches. The first approach used information derived from paleo-Köppen climate bands from @scoteseAtlas2021. Köppen bands are estimated from different fossil and lithological indicators and give a broad categorisation of spatial variation in major climate types (tropical, sub-tropical arid, warm temperate, cool temperate, and polar). We obtained estimated Köppen climate bands at 5 Ma intervals from 40 Ma to 0 Ma from and used the sub-tropical arid Köppen band as a binary aridity index. We removed sites found in the ocean by using digital elevation models (DEMs) to mask sites below sea-level. To account for uncertainty in paleogeographic reconstructions of continental movements and sea-levels, we used two alterative DEMs [@straumeGlobSed2019;@scotesePALEOMAP2018] to mask the ocean sites and provide two estimates of paleoaridity. To estimate paleotemperature from the Köppen bands, we followed the method of @scoteseAtlas2021 as implemented by @hagenEarth2021. Here, each Köppen climate band was assigned a temperature value based on present-day averages of those climate bands (tropical = 26, sub-tropical arid = 22, warm temperate = 16, cool temperate = 5, polar = -20). To account for global fluctuations in temperature, we then modified these values by a global temperature curve estimate from oxygen isotope data (66). Finally, to account for change of temperature with elevation, we used a lapse rate of 9.8°C/km. Following @hagenEarth2021, we model variation in the intensity of temperature oscillations during the Pleistocene. We also used the two alternative DEMs to give two alternative temperature curves using this approach [@straumeGlobSed2019;@scotesePALEOMAP2018].

The second approach used a global circulation model (GCM) to obtain a model-based estimate of paleotemperature and paleoprecipitation @valdesDeep2021. The paleoclimatic data is based on a HadCM3L model which couples ocean, atmosphere and vegetation and simulated at 5 Ma intervals based on the paleogeography of @scoteseAtlas2021. To make it comparable with the binary aridity index derived from the Köppen band approach, we converted the paleoprecipitation into an aridity index by thresholding the precipitation data. We applied a threshold of 500mm/year, below which is considered arid, above which was considered non-arid. Each of the three global paleoaridity and paleotemperature maps, from the two approaches, were projected into a Behrmann equal-area projection using a spatial resolution of 110km x 110km. We then temporally interpolated the data from 5 Ma intervals to 100 ka using bilinear interpolation. Finally, we subset the data to include only the continent of Sahul from 40 Ma to the present-day. From these maps, we obtain multiple Sahulian curves for paleotemperature by taking the mean, minimum and maximum values within Sahul at each timestep. For paleoaridity, we took the mean value, as minimum and maximum showed little variation in the binary metric.

@@

To evaluate how diversification rates correlated with climate and aridity through time, we fitted several time- and environemnt-dependent likelihood diversification models using the fit_bd and fit_env functions in the R package RPANDA [@morlonRPANDA2016] to the _Anilios_ phylogeny. Specifically, we fitted different combinations of pure-birth and birth-death models with speciation and/or extinction rates being constant or otherwise varying linearly or exponentially with paleoenvironment as described above. We fitted a total of 12 fit_env models based on reconstructed paleoenvironmental data to the phylogeny including three paleotemperature (mean, minimum, and maximum) and the paleoaridity index from each DEM. We also fit the phylogeny to the widely use mean global paleotemperature data set [@condamineMacroevolutionary2013] to provide comparison to our reconstructed paleoenvironment. To account for missing taxa, we used the fractioning function (f = 38/50) in RPANDA. We then assessed which diversification model best fit the data based on AICc scores.

@@

The four SSE models differed in whether they allowed diversification rates to be dependent on geographic states (range-dependent) or independent of them (range-independent), and whether they allowed an unmeasured trait ('hidden states') to influence diversification rates [@beaulieuDetecting2016;@caetanoHidden2018]. 
Model 1 assumed equal diversification rates regardless of geographic state. Model 2 allowed geographic range to affect diversification rates.
Models 3 and 4 were based on models 1 and 2, but they also allowed  unmeasured trait ('hidden state') to contribute to diversification rates. By incorporating hidden states, we could assess whether geographic range or an unmeasured trait contributed more to the observed diversification. For each model, we assumed uniform taxon sampling for 89% of arid lineages, 71% of non-arid lineages, and 83% of widespread lineages, based on the number of species in the phylogeny and all currently described species of _Anilios_. We then calculated AIC to identify the best-fitting models. Following the approach in @caetanoHidden2018, we averaged the estimates of diversification parameters across all models according to their AIC weights.

@@

The largest source of outward dispersal is the tropical grassland biome (16.2 ± 4.9 events), which makes up  43.6% compared outward dispersals from the arid biome (9.7 ± 3.2 events) of 26% (Fig. \@ref(fig:fig-biogeobears)c-d). Together the tropical grassland biome and the arid biome contributes the largest source of diversification accounting for 69.6% of outward transitions. On the other hand, the largest sink for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the Mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the Mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c-d).

Based on current distributions from ALA, most _Anilios_ species are found entirely or at least partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and Mediterranean biome (n = 10). We identified divergent biogeographic transitions between two major groups of species, which we will refer to as *Group 1* (n = 21 lineages, not including _A. proximus_) and *Group 2* (n = 14 lineages, not including _A. silvia_) (Fig. \@ref(fig:fig-biogeobears)a). The common ancestor of *Group 1* remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of *Group 2* first transitioned into the arid biome, and then 7/14 lineages expanded to the Mediterranean and temperate forest biomes, and/or transitioned back into the arid biome.

@@

Paleotemperature estimates for Australia differed between the three different methods (Fig. \@ref(fig:fig-paleoclimate-compare)b). However, all three paleotemperature models indicate Australia was hotter than the average global temperature (Fig. \@ref(fig:fig-paleoclimate-compare)a). Paleoclimate in Australia were also relatively constant compared to global temperature until the Plio-Pleistocene, a period characterised by fluctuations corresponding to the glacial cycles. 
Reconstructions of paleoaridity differed slightly between the two Köppen band methods, but these differed substantially to the aridity index derived from the GCM data (Fig. \@ref(fig:fig-paleoclimate-compare)b). Both Köppen band-based aridity indices show increasing aridity from 10 Ma, consistent with the onset of aridification of the continent [@byrneBirth2008;@pepperLife2021].
